Probabilistic Catalogs from Crowded Star Fields

### 2012-10-08 paper/ms.tex
Rename x1, x2 --> h1, h2

\usepackage[normalem]{ulem}  % For strike through...

@@

\newcommand{\etc}{\emph{etc.}}

@@

However, standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, uncertainty should be
taken into account. Instead of simply {\it estimating} the position and flux of
each object in the image, we should fully describe the fact that sometimes we
are not certain of the position and flux of each source. {\bf is it worth
commenting on the huge degeneracies and how poorly defined the objective is
unless we use prior knowledge?}

Essentially, the creation of a catalog is an attempt to answer the question
``given the image we have obtained, what objects are present in the field and
what are their properties?''. This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single answer to this question (i.e.~a single catalog), but rather a
probability distribution over the space of possible catalogs.
Constructing such a probability distribution is
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image that should be included in the catalog is itself unknown. Secondly, if
$N$ is large, then the parameter space
of positions and {\bf (physical?)} properties (flux, size, \etc) of the
objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties --- they
may take a long time to converge to the target posterior distribution over
catalogs.

@@

\section{Bayesian Inference}
To quantitatively model uncertainties, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta) \quad.
\end{equation}
\sout{Note that this is a very concise notation \citep{hogg}.}
{\bf I think that I know what you're trying to say here but I think that this
comment is more confusing than helpful...}
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}:
\begin{equation}
p(x|\theta) \quad.
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the data space, one for each possible value
of $\theta$. Note that
the sampling distribution is also an assertion of a prior state of knowledge
{\bf (or assumption?)}:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

@@

\section{Broken Power-Law Distribution}\label{power_law}
The broken power-law distribution is based on a straightforward extension
to a simple power-law distribution (also known as a Pareto distribution,
particularly in the statistics literature).
The broken power-law distribution for a variable $x$ is defined by:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcr}
0, & & x < x_1 \\
x^{-\alpha_1 - 1} & & x_1 \leq x \leq x_2 \\
x^{-\alpha_2 - 1} & & x > x_2
\end{array}
\right.
.
\end{eqnarray}
The free parameters of the broken power-law are:
\begin{eqnarray}
\beta = \{x_1, x_2, \alpha_1, \alpha_2\}
\end{eqnarray}

@@

With normalising terms included, the proportionality becomes an equality:
\begin{eqnarray}
p(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < x_1 \\
Z_1^{-1}x^{-\alpha_1 - 1}, & & x_1 \leq x \leq x_2 \\
Z_2^{-1}x^{-\alpha_2 - 1}, & & x > x_2
\end{array}
\right.
.
\end{eqnarray}

Two conditions will be used to determine the normalisers $Z_1$ and $Z_2$.
Firstly, the PDF should be continuous at $x=x_2$:
\begin{eqnarray}
Z_1^{-1}x_2^{-\alpha_1 - 1} &=& Z_2^{-1}x_2^{-\alpha_2 - 1}\\
\implies
Z_2 &=& Z_1x_2^{\alpha_1-\alpha_2}
\end{eqnarray}

@@

\begin{eqnarray}
\int_{x_1}^{x_2} Z_1^{-1} x^{-\alpha_1 - 1} \, dx
+
\int_{x_2}^\infty Z_2^{-1} x^{-\alpha_2 - 1} \, dx
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[x_1^{-\alpha_1} - x_2^{-\alpha_1}\right]
+
Z_2^{-1}\alpha_2^{-1}x_2^{-\alpha_2}
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[x_1^{-\alpha_1} - x_2^{-\alpha_1}\right]
+
Z_1^{-1}x_2^{\alpha_2-\alpha_1}\alpha_2^{-1}x_2^{-\alpha_2}
&=& 1
\end{eqnarray}

\begin{eqnarray}
\implies
Z_1 &=& \alpha_1^{-1}\left[x_1^{-\alpha_1} - x_2^{-\alpha_1}\right]
+
x_2^{-\alpha_1}\alpha_2^{-1}
\end{eqnarray}

The cumulative distribution (CDF) is:
\begin{eqnarray}
P(X \leq x) = F(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < x_1 \\
(Z_1\alpha_1)^{-1}\left(x_1^{-\alpha_1} - x^{-\alpha_1}\right), & & x_1 \leq x \leq x_2 \\
1 - (Z_2\alpha_2)^{-1}x^{-\alpha_2}, & & x > x_2
\end{array}
\right.
\end{eqnarray}
and the inverse CDF is:
\begin{eqnarray}
F^{-1}(u) &=&
\left\{
\begin{array}{lcr}
\left[x_1^{-\alpha_1} - uZ_1\alpha_1\right]^{-1/\alpha_1}, & & 0 < u < 1 - (Z_2\alpha_2)^{-1}x_2^{-\alpha_2}\\
\left[Z_2\alpha_2(1-u)\right]^{-1/\alpha_2},& & 1 - (Z_2\alpha_2)^{-1}x_2^{-\alpha_2}< u < 1
\end{array}
\right.
\end{eqnarray}

>>>>

@@

@@

However, standard methods for constructing catalogs can have difficulty in some
challenging situations. For example, when multiple sources overlap partially
or completely, it can be difficult to determine how many sources are present,
and how much flux belongs to each source. In principle, uncertainty should be
taken into account. Instead of simply {\it estimating} the position and flux of
each object in the image, we should fully describe the fact that sometimes we
are not certain of the position and flux of each source.

Essentially, the creation of a catalog is an attempt to answer the question
``given the image we have obtained, what objects are present in the field and
what are their properties?''. This motivates a probabilistic (Bayesian)
approach to making catalogs. The output of such an approach would not be a
single answer to this question (i.e. a single catalog), but rather a
probability distribution over the space of possible catalogs.
Constructing such a probability distribution is
challenging for a number of reasons. Firstly, the number $N$ of objects in the
image (and that should be in the catalog) is itself unknown. Secondly, if $N$ is large, then the parameter space
of positions and properties (flux, size, etc) of the objects is also large.
This can cause Markov Chain Monte Carlo (MCMC) algorithms difficulties -- they
may take a long time to converge to the target posterior distribution over
catalogs.

@@

\section{Bayesian Inference}
To quantitatively model uncertainties, Bayesian Inference is the appropriate
framework \citep{cox, jaynes, caticha}. Suppose there exist unknown parameters
(denoted
collectively by $\theta$) and we expect to obtain some data $x$. Our prior
state of knowledge about the parameters is modelled by a prior
probability distribution:
\begin{equation}
p(\theta)
\end{equation}.
Note that this is a very concise notation \citep{hogg}.
We also model how the parameters give rise to the data, via a generative model.
This is also known as a {\it sampling distribution}.
\begin{equation}
p(x|\theta)
\end{equation}
Despite the singular, the sampling distribution is actually a family of
probability distributions over the data space, one for each possible value
of $\theta$. Note that
the sampling distribution is also an assertion of a prior state of knowledge:
It models prior information about the fact that the data $x$ is connected to
the parameters $\theta$ in some way. Without this prior knowledge, learning is
impossible: there has to be some relationship between the parameters and the
data, otherwise new data tells you about nothing but itself.

@@

\section{Broken Power-Law Distribution}\label{power_law}
The broken power-law distribution is based on a straightforward extension
to a simple power-law distribution (also known as a Pareto distribution,
particularly in the statistics literature). The power-law distribution for a
variable $x$ (given a lower cutoff $x=h$ and a slope $\alpha$) is defined by:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcr}
0, & & x < h \\
x^{-\alpha - 1} & & x \geq h.
\end{array}
\right.
\end{eqnarray}
In contrast, the broken power-law distribution for a variable $x$ is defined by
a lower cutoff $x=h_1$, two slopes $\{\alpha_1, \alpha_2\}$ and a break point
$x=h_2$:
\begin{eqnarray}
p(x) &\propto&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
x^{-\alpha_1 - 1} & & h_1 \leq x \leq h_2 \\
x^{-\alpha_2 - 1} & & x > h_2.
\end{array}
\right.
\end{eqnarray}
The free parameters of the broken power-law are:
\begin{eqnarray}
\beta = \{h_1, h_2, \alpha_1, \alpha_2\}
\end{eqnarray}

@@

With normalising terms included, the proportionality becomes an equality:
\begin{eqnarray}
p(x) &=&
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
Z_1^{-1}x^{-\alpha_1 - 1}, & & h_1 \leq x \leq h_2 \\
Z_2^{-1}x^{-\alpha_2 - 1}, & & x > h_2
\end{array}
\right.
.
\end{eqnarray}

Two conditions will be used to determine the normalisers $Z_1$ and $Z_2$.
Firstly, the PDF should be continuous at $x=h_2$:
\begin{eqnarray}
Z_1^{-1}h_2^{-\alpha_1 - 1} &=& Z_2^{-1}h_2^{-\alpha_2 - 1}\\
\implies
Z_2 &=& Z_1h_2^{\alpha_1-\alpha_2}
\end{eqnarray}

@@

\begin{eqnarray}
\int_{h_1}^{h_2} Z_1^{-1} x^{-\alpha_1 - 1} \, dx
+
\int_{h_2}^\infty Z_2^{-1} x^{-\alpha_2 - 1} \, dx
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_2^{-1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1 \\
Z_1^{-1}\alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
Z_1^{-1}h_2^{\alpha_2-\alpha_1}\alpha_2^{-1}h_2^{-\alpha_2}
&=& 1
\end{eqnarray}

\begin{eqnarray}
\implies
Z_1 &=& \alpha_1^{-1}\left[h_1^{-\alpha_1} - h_2^{-\alpha_1}\right]
+
h_2^{-\alpha_1}\alpha_2^{-1}
\end{eqnarray}

The cumulative distribution (CDF) is:
\begin{eqnarray}
P(X \leq x) = F(x) &=& 
\left\{
\begin{array}{lcr}
0, & & x < h_1 \\
(Z_1\alpha_1)^{-1}\left(h_1^{-\alpha_1} - x^{-\alpha_1}\right), & & h_1 \leq x \leq h_2 \\
1 - (Z_2\alpha_2)^{-1}x^{-\alpha_2}, & & x > h_2 
\end{array}
\right.
\end{eqnarray}
and the inverse CDF is:
\begin{eqnarray}
F^{-1}(u) &=& 
\left\{
\begin{array}{lcr}
\left[h_1^{-\alpha_1} - uZ_1\alpha_1\right]^{-1/\alpha_1}, & & 0 < u < 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}\\
\left[Z_2\alpha_2(1-u)\right]^{-1/\alpha_2},& & 1 - (Z_2\alpha_2)^{-1}h_2^{-\alpha_2}< u < 1
\end{array}
\right.
\end{eqnarray}
